Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-07-08 manuscript/manuscript.Rmd
minor edits, suggestions, and questions added

---
title: Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes
  and Reservoirs
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the
  focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts
  of eutrophication, but to determine the effectiveness requires long-term datasets.  A
  recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE),
  found stable water quality in the Northeast and Midwestern regions of the United
  States; however, trends at smaller scales may be masked.  We address this by analyzing
  the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW)
  dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs
  for over 25 years. These data, included in LAGOSNE, allow for comparison of water
  quality trends at regional and state extents. We assess for trends with z-scores
  (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature
  and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing
  trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total
  z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll
  is increasing, in spite of stable nutrient regimes. Although not causal, this analysis
  suggests an association between lake temperature and productivity.  Additionally,
  we demonstrate both the value of long-term monitoring programs, like URIWW, for
  identifying trends in environmental condition, and the utility of site-specific
  z-scores for analyzing for long-term trends. '
  nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific
  *a*, TN, TP, and TN:TP.  In short, productivity in Rhode Island lakes and reservoirs
---

@@

Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global]. Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes]. Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].  This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate]. To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication]. These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.  Let's discuss it). Maybe...  These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.(I think the edited sentence improves and clarifies the original)

Attempts to explore these complex spatial dynamics are ongoing. (my attempt at intro sentence for paragraph).  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.  Again, let's chat). Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (i.e. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long]. (thought- Maybe this paragraph just needs a introduction sentence. )

Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper)(when used as adjectives, regional descriptors may or may not be capitalized depending on style (i.e. Chicago keeps lowercase) but should be capitalized when describing the region) United States have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e. subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the Northeastern and Midwestern US [@filstrup2014regional; @filstrup2018evidence]. Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.

@@

Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.  First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states? Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends. Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).  The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years. The program began in 1988, monitoring 14 lakes, and has now grown to include over 250 monitoring sites on over 120 waterbodies, including lakes, reservoir, rivers, streams, and estuaries with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data (^I don't know what this means- is a common knowledge I'm missing? If so, ignore), and is an integral part of the state's environmental data collection strategy. Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field and are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory.

@@

In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less. For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, and `r length(total_p_10yr_sites)` samples for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

@@

1. For each site, calculate the annual means, producing a single mean value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site means across years are assumed to be independent.
2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g. $\beta$~1~ = 0).

@@

Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].  While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.  The analysis conducted by Oliver et al. [-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al. [-@Oliver_2017] but using the trend analysis approach outlined above.

@@

During the period of 1990 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

@@

Average annual scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).  Panel A. Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).

@@

```{r no_early_tn, echo=FALSE}
ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```

Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score). Panel A. Yearly averaged TN z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).  TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).  Panel A. Yearly averaged TP z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TP z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).  Panel A. Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].

@@

In general, there was little evidence to suggest broad regional changes. Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)

>>>>

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication but determining the effectiveness requires long-term datasets.  A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE) found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at smaller scales may be masked.  We address a smaller scale by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents. We assess for trends with z-scores (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP.  In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes. Although not causal, this analysis suggests an association between lake temperature and productivity.  Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends. '
---

@@

Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global]. Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes]. Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].  This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate]. To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication]. These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.  Let's discuss it). Maybe...  These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales. (I agree with the previous comment and like the suggested sentence)

Attempts to explore these spatial dynamics are ongoing (my attempt at intro sentence for paragraph).  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.  Again, let's chat). Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (i.e. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long]. (thought- Maybe this paragraph just needs a introduction sentence. )

Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper)(It is my understanding that when using as a title, these should be capitalized, as compass directions not capitalized) United States have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e. subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the Northeastern and Midwestern US [@filstrup2014regional; @filstrup2018evidence]. Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.

@@

Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.  First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United States? Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends. Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).  The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years. The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data (^I don't know what this means- is a common knowledge I'm missing? If so, ignore), and is an integral part of the state's environmental data collection strategy. Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory. (I am not familiar with this certification. Is there a citation or certificate # to support?)

@@

In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 (Throughout the methods and discussion this varies between 1990 and 1993. It would be good to be consistent) and 2016, 2) were sampled between May and October, 3) and were sampled at a depth of 2 meters or less. For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] (This is repeated from the intro).  We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

@@

1. For each site, calculate the annual means producing a single mean value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site means across years are assumed to be independent.
2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g. $\beta$~1~ = 0). (This renders strangely)

@@

Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].(This is repeated from the intro)  While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.  The analysis conducted by Oliver et al. [-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al. [-@Oliver_2017] but using the trend analysis approach outlined above.

@@

During the period of 1990 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`). (add degrees to celsius?)

@@

Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`), and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).  Panel A. Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).

@@

```{r no_early_tn, echo=FALSE}
ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```
(Will the figures be reproduced in color for the print version? If not, should the captions be changed to not indicate colors?)
Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score). Panel A. Yearly averaged TN z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Since 1995, the yearly trend shows a much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).  TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) (The slope and p-values are not the same as the figure) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).  Panel A. Yearly averaged TP z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TP z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).  Panel A. Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].

@@

In general, there was little evidence to suggest broad regional changes. Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio showed little change (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
